---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gfwr

<!-- badges: start -->
<!-- badges: end -->

The `gfwr` R package is a simple wrapper for the Global Fishing Watch (GFW) APIs. It provides convenient functions to pull GFW data directly into R in tidy formats.

...more

## Installation

You can install the development version of gfwr like so:

``` r
devtools::install_github("GlobalFishingWatch/gfwr")
```

## Authorization

The use of `gfwr` requires a GFW API token, which users can request from the [GFW API Portal](). Save this token to your `.Renviron` file as a string called `"GFW_TOKEN"`. Then use `gfw_auth()` to extract the token and pass it to subsequent `gfwr` functions.

## Example

This is a basic example which shows you how to solve a common problem:

```{r example, eval=F}
library(gfwr)

# Get API key from local .Renviron file
key <- gfw_auth()
```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.


