---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gfwr

<!-- badges: start -->
<!-- badges: end -->

<!-- Add link to API documentation page once it's ready-->

The `gfwr` R package is a simple wrapper for the Global Fishing Watch (GFW) [APIs](https://api-doc.dev.globalfishingwatch.org/#introduction). 
It provides convenient functions to pull GFW data directly into R in tidy formats.

The package currently works with the following APIs:

* Vessel API: vessel search and identity based on AIS self reported data
* Events API: encounters, loitering, port visits and fishing events based on AIS data
* Map Visualization (4Wings API): apparent fishing effort based on AIS data


## Installation

You can install the development version of gfwr like so:

``` r
devtools::install_github("GlobalFishingWatch/gfwr")
```

Once everything is installed, you can load and use `gfwr` 
in your scripts with `library(gfwr)`

```{r, eval = FALSE}
library(gfwr)
```

```{r, eval = TRUE, echo = FALSE, message = FALSE}
devtools::load_all()
```

## Authorization

The use of `gfwr` requires a GFW API token, which users can request from
the [GFW API Portal](https://globalfishingwatch.org/ocean-engine/tokens/signup). Save
this token to your `.Renviron` file (using `usethis::edit_r_environ()`) by adding a string
named `"GFW_TOKEN"` to the file (`"GFW_TOKEN" = "PASTE_YOUR_TOKEN_HERE"`). Save 
the `.Renviron` file and restart the R session to make the edit effective.  

Then use the `gfw_auth` helper function to save the information in an object
in your R workspace every time you need to extract the token and pass it
to subsequent `gfwr` functions.

So you can do:

```{r, eval = T}
key <- gfw_auth()

# or this
key <- Sys.getenv("GFW_TOKEN")
```


## Vessels API

The `get_vessel_info` function allows you to get vessel identity details. 
There are three search types: `basic`, `advanced`, and `id`.

* `basic` search takes features like MMSI, IMO, callsign, shipname as inputs and 
identifies all vessels in the specified dataset that match
* `advanced` search allows for the use of fuzzy matching with terms such as LIKE. The `id` search allows the user to search using a GFW vessel
* `id` allows the user to specify the `vessel id` (generated by GFW)

> **Note**: 
<br>
> `vessel id` is an internal ID generated by GFW to connect data accross APIs 
and involves a combination of vessel and tracking data information

The user can also specify which identity databases to use: `carrier_vessel`, 
`support_vessel`, `fishing_vessel`, or `all`. With the latter, all databases are used for the search. 
This is generally recommended and is the option set by default. 


### Examples

To get information of a vessel with MMSI = 224224000 using all datasets:

```{r example_vessel_info_1, eval=T}
get_vessel_info(query = 224224000, search_type = "basic", 
                dataset = "all", key = key)
```

To combine different fields and do fuzzy matching to search the `carrier vessel`
dataset:

```{r example_vessel_info_2, eval=F}
get_vessel_info(query = "shipname LIKE '%GABU REEFE%' OR imo = '8300949'", 
                search_type = "advanced", dataset = "carrier_vessel", key = key)
```

To specify a `vessel id`:

```{r example_vessel_info_3, eval=F}
get_vessel_info(query = "8c7304226-6c71-edbe-0b63-c246734b3c01", 
                search_type = "id", dataset = "carrier_vessel", key = key)
```

To specify more than one `vessel id`:
> **Note**: 
<br>
> No spaces or newlines are permitted between the `vessel ids` 

```{r example_vessel_info_4, eval=F}
get_vessel_info(query = 
                  "8c7304226-6c71-edbe-0b63-c246734b3c01,6583c51e3-3626-5638-866a-f47c3bc7ef7c,71e7da672-2451-17da-b239-857831602eca", 
                search_type = 'id', key = key)
```


## Events API

The `get_event` function allows you to get data from vessel activities such as: 
apparent fishing events, encounters, loitering, and port visits. 
Find more information in our [caveat documentation](https://api-doc.dev.globalfishingwatch.org/#data-caveat).



<!-- I don't think we have tested loitering, or encounters yet-->


<!-- #' Base function to get event from API and convert response to data frame -->
<!-- #' -->
<!-- #' @param event_type Type of event to get data of. It can be "port_visit" or "fishing" -->
<!-- #' @param vessel VesselID. How to get this? -->
<!-- #' @param include_regions Whether to include regions? Ask engineering if this can always be false -->
<!-- #' @param start_date Start of date range to search events -->
<!-- #' @param end_date End of date range to search events -->
<!-- #' @param key Authorization token. Can be obtained with gfw_auth function -->

### Examples

Let's say that you don't know the `vessel id` but you have the MMSI (or other 
identity information). 
You can use `get_vessel_info` function first to extract `vessel id` and then use
it in the `get_event` function:

```{r example_id_event, eval=T}
vessel_id <- get_vessel_info(query = 224224000, search_type = "basic", key = key)$id
```

To get a list of port visits for that vessel:

```{r example_event_type_1, eval=T}
get_event(event_type='port_visit',
                      vessel = vessel_id,
                      key = key
                      )
```

We can also use more than one `vessel id`:

```{r example_event_type_2, eval=F}
get_event(event_type='port_visit',
                      vessel = '8c7304226-6c71-edbe-0b63-c246734b3c01,6583c51e3-3626-5638-866a-f47c3bc7ef7c',
                      include_regions = NULL,
                      start_date = "2020-01-01",
                      end_date = "2020-02-01",
                      key = key
                      )
```

Or get encounters for all vessels in a given date range:

```{r example_event_type_3, eval=F}
get_event(event_type='encounter',
                      start_date = "2020-01-01",
                      end_date = "2020-02-01",
                      key = key
                      )
```



## Map Visualization API 

The `get_raster` function gets a raster from the 4Wings API and converts the 
response to a data frame. In order to use it, you should specify:

* The spatial resolution, which can be `low` (0.1 degree) or `high` (0.01 degree)
* The temporal resolution, which can be `daily`, `monthly`, or `yearly`.
* The variable to group by: `vessel_id`, `flag`, `geartype`, or `flagAndGearType`
* The date range 
* The `geojson` region or region code (such as an EEZ code) to filter the raster
* The source for the specified region (currently, `eez`, `mpa`, or `user_json`)


### Examples

Here's an example where we enter the geojson data manually:


```{r example_map_1, eval=F}

region_json = '{"geojson":{"type":"Polygon","coordinates":[[[-76.11328125,-26.273714024406416],[-76.201171875,-26.980828590472093],[-76.376953125,-27.527758206861883],[-76.81640625,-28.30438068296276],[-77.255859375,-28.767659105691244],[-77.87109375,-29.152161283318918],[-78.486328125,-29.45873118535532],[-79.189453125,-29.61167011519739],[-79.892578125,-29.6880527498568],[-80.595703125,-29.61167011519739],[-81.5625,-29.382175075145277],[-82.177734375,-29.07537517955835],[-82.705078125,-28.6905876542507],[-83.232421875,-28.071980301779845],[-83.49609375,-27.683528083787756],[-83.759765625,-26.980828590472093],[-83.84765625,-26.35249785815401],[-83.759765625,-25.64152637306576],[-83.583984375,-25.16517336866393],[-83.232421875,-24.447149589730827],[-82.705078125,-23.966175871265037],[-82.177734375,-23.483400654325635],[-81.5625,-23.241346102386117],[-80.859375,-22.998851594142906],[-80.15625,-22.917922936146027],[-79.453125,-22.998851594142906],[-78.662109375,-23.1605633090483],[-78.134765625,-23.40276490540795],[-77.431640625,-23.885837699861995],[-76.9921875,-24.28702686537642],[-76.552734375,-24.846565348219727],[-76.2890625,-25.48295117535531],[-76.11328125,-26.273714024406416]]]}}'

get_raster(spatial_resolution = 'low',
           temporal_resolution = 'yearly',
           group_by = 'flag',
           date_range = '2020-01-01,2021-10-01',
           region = region_json,
           region_source = 'user_json',
           key = key)

```

If you want raster data from a particular EEZ, you can use the `get_region_id` 
function to get the EEZ id, enter that code in the `region` argument
of `get_raster` instead of the geojson data (ensuring you specify the `region_source`
as `'eez'`:

```{r example_map_2, eval=F}

# use EEZ function to get EEZ code of Cote d'Ivoire
code_eez <- get_region_id(region_name = 'CIV', region_source = 'eez', key = key)

get_raster(spatial_resolution = 'low',
           temporal_resolution = 'yearly',
           group_by = 'flag',
           date_range = '2021-01-01,2021-10-01',
           region = code_eez$id,
           region_source = 'eez',
           key = key)
```

You could search for just one word in the name of the EEZ and then decide which 
one you want:

```{r example_map_3, eval=T}

(get_region_id(region_name = 'French', region_source = 'eez', key = key))

# Let's say we're interested in the French Exclusive Economic Zone, 5677
get_raster(spatial_resolution = 'low',
           temporal_resolution = 'yearly',
           group_by = 'flag',
           date_range = '2021-01-01,2021-10-01',
           region = 5677,
           region_source = 'eez',
           key = key)
```

A similar approach can be used to search for a specific Marine Protected Area, 
in this case the Phoenix Island Protected Area (PIPA)

```{r example_map_4, eval=F}

# use region id function to get MPA code of Phoenix Island Protected Area
code_mpa <- get_region_id(region_name = 'Phoenix', region_source = 'mpa', key = key)

get_raster(spatial_resolution = 'low',
           temporal_resolution = 'yearly',
           group_by = 'flag',
           date_range = '2015-01-01,2018-01-01',
           region = code_mpa$id[1],
           region_source = 'mpa',
           key = key)
```

It is also possible to filter rasters to one of the five regional fisheries management
organizations (RFMO) that manage tuna and tuna-like species. These include `"ICCAT"`,
`"IATTC"`,`"IOTC"`, `"CCSBT"` and `"WCPFC"`.

```{r example_map_5, eval=F}
get_raster(spatial_resolution = 'low',
           temporal_resolution = 'daily',
           group_by = 'flag',
           date_range = '2021-01-01,2021-01-15',
           region = 'ICCAT',
           region_source = 'trfmo',
           key = key)
```

