---
title: "Using gfwr: identity and vessel ID fields"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using gfwr: identity and vessel if fields}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = F}
library(gfwr)
```

`gfwr` can be used to access the vessel database Global Fishing Watch has compiled and maintained over the years.
On our map, the vessel search is located at https://globalfishingwatch.org/map/vessel-search.

Function `get_vessel_info()` provides the same functionality as the map, it can search for vessels using any of the identifiers below:

- MMSI (Maritime Mobile Service Identity) are 9-digit numbers broadcasted in AIS messages. They are supposed to be unique for each vessel but can be duplicated for many reasons--including the fact that the data entry is manual.
- IMO numbers are permanent unique identifiers that follow a vessel from construction to scrapping. Given by the International Maritime Organization
- Callsign

## Registry data 


Due to the complexity of this identity markers, Global Fishing watch created an internal vesselID to summarize the findings about vessels identity markers. 
While MMSI is still one of the main 


Search by any 

Search by more complex combinations 

Park et al 2023 and the need for vessel id

other apis use vessel id

# Getting vessel id

# Read more:

- https://globalfishingwatch.org/datasets-and-code-vessel-identity/
- https://globalfishingwatch.org/data/spoofing-one-identity-shared-by-multiple-vessels/ 

