---
title: "Making maps with `gfwr`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{making-maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6
)
```

A powerful feature of `gfwr` is the ability to get data for making custom maps of apparent fishing effort and vessel activities. This vignette demonstrates how to combine multiple `gfwr` functions with `ggplot2` and `sf` to make a variety of maps of fishing vessel activity.

Specifically, this vignette will show how to make maps of:

-   Global apparent fishing effort
-   Apparent fishing effort by a specific flag state
-   Apparent fishing effort within a specific Exclusive Economic Zone (EEZ)
-   Encounter events within a regional fisheries management organization (RFMO)
-   Activity within a custom GeoJSON

First, load `gfwr`.

```{r setup}
library(gfwr)
```

### Other packages

For this vignette, we'll also use some additional packages for plotting (`ggplot2`,`sf`) and reference data (`rnaturalearth`)

```{r other_packages}
library(tidyverse)
library(sf)
library(rnaturalearth)
```

### Plot theme, palletes, and reference layers

```{r}
# Map theme with dark background
map_theme <- ggplot2::theme_minimal() + 
  ggplot2::theme(
    panel.border = element_blank(), 
    panel.background = element_rect(fill = "#0a1738", color = NA), 
    legend.position = "bottom", legend.box = "vertical", 
    legend.key.height = unit(3, "mm"), 
    legend.key.width = unit(20, "mm"), 
    legend.title.align = 0.5, 
    legend.text = element_text(color = "#848b9b", size = 8), 
    legend.title = element_text(face = "bold", color = "#363c4c", size = 8), 
    plot.title = element_text(face = "bold", color = "#363c4c", size = 10), 
    plot.subtitle = element_text(color = "#363c4c", size = 10), 
    axis.title = element_blank(), 
    axis.text = element_text(color = "#848b9b", size = 6), 
    panel.grid.major = element_line(color = "#0a1738"), 
    panel.grid.minor = element_line(color = "#0a1738"))

# Palette for fishing activity
map_effort_dark <- c("#0c276c", "#3b9088", "#eeff00", "#ffffff")
```

## Global apparent fishing effort

## Activity by EEZ

To get activity for a specific EEZ, we first use the `get_region_id()` function to find the `id` of our EEZ of interest using the EEZ ISO3 code. Note that, for some countries, the ISO3 code will return multiple regions

```{r}
# use EEZ function to get EEZ code of Cote d'Ivoire
code_eez <- get_region_id(region_name = 'ITA', region_source = 'eez', key = gfw_auth())

eez_fish_df <- get_raster(spatial_resolution = 'low',
           temporal_resolution = 'yearly',
           group_by = 'flag',
           date_range = '2021-01-01,2021-10-01',
           region = code_eez$id,
           region_source = 'eez')
```

The data includes fishing by all flag states, so to make a map of all activity, we first need to summarize activity by grid cell.

```{r}
eez_fish_all_df <- eez_fish_df %>% 
  group_by(Lat, Lon) %>% 
  summarize(fishing_hours = sum(`Apparent Fishing hours`, na.rm = T))
```

Now we can use `ggplot2` to plot the data:

```{r}
eez_fish_all_df %>% 
  ggplot() +
  geom_raster(aes(x = Lon,
                  y = Lat,
                  fill = fishing_hours)) +
  geom_sf(data = ne_countries(returnclass = 'sf', scale = 'medium')) +
  coord_sf(xlim = c(min(eez_fish_all_df$Lon),max(eez_fish_all_df$Lon)),
           ylim = c(min(eez_fish_all_df$Lat),max(eez_fish_all_df$Lat))) +
  scale_fill_gradientn(colors = map_effort_dark, na.value = NA) +
  map_theme
```

The `get_region_id` function also works in reverse. If a region id is passed as a `numeric` to the function as the `region_name`, the corresponding region label or iso3 can be returned. This is especially useful when events are returned with regions.

```{r example_region_id, eval=T}
# using same example as above
encounters_df <- get_event(event_type = 'encounter',
          start_date = "2021-01-01",
          end_date = "2021-10-01",
          include_regions = TRUE) 

encounters_hs_df <- encounters_df %>%
  # extract EEZ id code
  dplyr::mutate(eez = as.character(purrr::map(purrr::map(regions, pluck, 'eez'), 
                                              paste0, collapse = ','))) %>%
  dplyr::select(id, type, start, end, lat, lon, eez) %>%
  dplyr::filter(eez == '') 
```

Encounters have two rows per event to represent both vessels. Extract the event ID and select one row per event.

```{r}
encounters_hs_sf_df <- encounters_hs_df %>% 
  separate(id, c('event_id','vessel_number')) %>% 
  filter(vessel_number == 1) %>% 
  st_as_sf(coords = c('lon','lat'), crs = 4326)
```

```{r}
ggplot() +
  geom_sf(data = ne_countries(returnclass = 'sf', scale = 'small')) +
  geom_sf(data = encounters_hs_sf_df, color = 'red', alpha = 0.4, size = 1) +
  map_theme
```

## Activity within a custom region

The `get_raster` function allows users to download fishing effort data within a custom region by providing a GeoJSON string object.

```{r}
region_json = '{"geojson":{"type":"Polygon","coordinates":[[[-15.1296846569,46.500565613],[13.2586868107,46.500565613],[13.2586868107,61.6189445123],[-15.1296846569,61.6189445123],[-15.1296846569,46.500565613]]]}}'

geojson_df <- get_raster(spatial_resolution = 'low',
           temporal_resolution = 'yearly',
           group_by = 'flag',
           date_range = '2021-01-01,2021-12-31',
           region = region_json,
           region_source = 'user_json')
```

```{r}
geojson_all_df <- geojson_df %>% 
  group_by(Lat, Lon) %>% 
  summarize(fishing_hours = sum(`Apparent Fishing hours`, na.rm = T))
```

```{r}
geojson_all_df %>% 
  ggplot() +
  geom_raster(aes(x = Lon,
                  y = Lat,
                  fill = fishing_hours)) +
  geom_sf(data = ne_countries(returnclass = 'sf', scale = 'medium')) +
  coord_sf(xlim = c(min(geojson_all_df$Lon),max(geojson_all_df$Lon)),
           ylim = c(min(geojson_all_df$Lat),max(geojson_all_df$Lat))) +
  scale_fill_gradientn(colors = map_effort_dark, na.value = NA) +
  map_theme
```
