---
title: "Exploring the vessel presence layer in `gfwr`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring the vessel presence layer in `gfwr`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = 'center',
  out.width = "90%"
  # fig.width = 6
)
```

## Overview

The `gfw_ais_presence()` function provides gridded vessel presence data from Global Fishing Watch's [4Wings Map Visualization API](https://globalfishingwatch.org/our-apis/documentation#map-visualization-4wings-api). 
Global Fishing Watch's vessel presence layer includes all vessel types, fishing 
or not, and summarizes their presence in hours, based on the data transmitted by 
their AIS transponders. 

In this vignette we will explore several ways to call this function, and the filters that have been implemented in our APIs to support the exploration of this layer. 


## Setup

To get started, first load the `gfwr` package and some of the packages we will use

```{r setup, eval = F}
library(gfwr)
```

```{r load, eval = T, echo = F}
devtools::load_all()
```


```{r other_packages, message = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(glue)
```

We will fetch data for January-March 2024. __Remember the date intervals in `gfwr` include the start of the interval and *exclude the last date* of the interval__:

```{r dates}
start_date <- '2024-01-01' # will be included
end_date <- '2024-04-01'   # will be excluded. search will be up to 2024-03-31
```



## Handling pre-built regions: MPAs, RFMOs, and EEZs


`gfw_ais_presence()` was designed to provide data for a specific region, offering users the ability to select from multiple built-in region types by specifying a specific Exclusive Economic Zone (EEZ), Marine Protected Area (MPA), or Regional Fisheries Management Organization (RFMO).

> __Note:__ The use of a region is mandatory, as the API is not designed to handle global requests


The list of available regions for each type, and their `label` and `id`, can be accessed with the `gfw_regions()` function.

```{r eez_region_codes}
eez_regions <- gfw_regions(region_source = 'EEZ')
eez_regions
```

`gfwr` also includes the `gfw_region_id()` function to get the `label` and `id` for a specific region using the `region` argument. 
For EEZs, `region` corresponds to the name or the country or the ISO3 code. Note that, for some countries, the name will return multiple regions. For RFMOs, `region` corresponds to the RFMO abbreviation (e.g. `"ICCAT"`) and for MPAs it refers to the name of the MPA. 

To fetch the numeric code of the Senegal EEZ, let's use `gfw_region_id()`

```{r code_eez}
# Use gfw_region_id function to get EEZ code for Senegal
senegal_eez_code <- gfw_region_id(region = "Senegal", region_source = "EEZ")
senegal_eez_code
```

The results show the EEZ and a Joint regime area. We will pick the EEZ code, `8371`. 


## Calling the function

The `gfw_ais_presence()` function allows users to specify multiple criteria to customize the data they download, including the date range, spatial and temporal resolution, and grouping variables. See the documentation for `gfw_ais_presence()` or the [GFW APIs](https://globalfishingwatch.org/our-apis/documentation#report-url-parameters-for-both-post-and-get-requests) for more info about these parameter options.

Spatial resolution can be LOW = 0.1 degree or HIGH = 0.01 degree, 

```{r}
vp_senegal <- gfw_ais_presence(spatial_resolution = "LOW",
                               temporal_resolution = "MONTHLY",
                               start_date = start_date,
                               end_date = end_date,
                               region_source = "EEZ",
                               region = 8371)
vp_senegal
```

Without grouping variables, the function will return a number of vessel IDs present (see our vessel identity vignette) and the total vessel presence hours for each cell (lat, lon). 

The `Time Range` column will be expressed in the temporal units of the temporal resolution selected. In this example, `MONTHLY` will create a Time Range expressed in months: `YYYY-MM`

Explore other temporal resolution and how the results vary.

```{r}
vp_senegal <- gfw_ais_presence(spatial_resolution = "LOW",
                               temporal_resolution = "YEARLY",
                               start_date = start_date,
                               end_date = end_date,
                               region_source = "EEZ",
                               region = 8371)
vp_senegal
```

## Grouping variables

The outputs of `gfw_ais_presence()` can be grouped by `FLAG`, `GEARTYPE`, `FLAGANDGEARTYPE`, `MMSI` or `VESSEL_ID`. 
This will create extra grouping columns, and the number of vessel presence hours will be expressed accordingly. 

```{r}
vp_senegal_flag <- gfw_ais_presence(spatial_resolution = "LOW",
                                    temporal_resolution = "MONTHLY",
                                    group_by = "FLAG",
                                    start_date = start_date,
                                    end_date = end_date,
                                    region_source = "EEZ",
                                    region = 8371)
vp_senegal_flag |> count(flag) |> arrange((desc(n)))

```

__Note__ that these results are grouped at the `Vessel ID` scale, which is not the same as the number of vessels. 
Check our [Vessel identity vignette](/articles/identity) for more information



Grouping by MMSI will group the results at the MMSI scale, which can correspond to individual vessels, but this is not always the case. 

```{r}
vp_senegal_MMSI <- gfw_ais_presence(spatial_resolution = "LOW",
                                    temporal_resolution = "MONTHLY",
                                    group_by = "MMSI",
                                    start_date = start_date,
                                    end_date = end_date,
                                    region_source = "EEZ",
                                    region = 8371)
vp_senegal_MMSI
```


Finally, grouping by `Vessel ID` returns all the details about the vessel identity of the vessels present in the area: 

```{r}
vp_senegal_vesselID <- gfw_ais_presence(spatial_resolution = "LOW",
                                        temporal_resolution = "MONTHLY",
                                        group_by = "VESSEL_ID",
                                        start_date = start_date,
                                        end_date = end_date,
                                        region_source = "EEZ",
                                        region = 8371)
vp_senegal_vesselID |> count(`Gear Type`)
vp_senegal_vesselID |> count(`Vessel Type`)
```

The columns include `r paste(names(vp_senegal_vesselID), collapse = ", ")`.
This is an important option to understand which vessels are active in an area, and their characteristics. 


<!--
## Filter options

### Speed classes 

`gfw_ais_vessel_presence()` supports filtering the vessels by speed range (in knots) in the following categories : 

- <2 – Less than 2 knots
- 2-4 – 2 to 4 knots
- 4-6 – 4 to 6 knots
- 6-10 – 6 to 10 knots
- 10-15 – 10 to 15 knots
- 15-25 – 15 to 25 knots
- \>25 – Greater than 25 knots
    
```{r gfw_speed, eval = FALSE}
eez_vessel_presence_speed <- gfw_ais_presence(
  spatial_resolution = "LOW",
  temporal_resolution = "YEARLY",
  group_by = "FLAG",
  filter_by = "speed = '<2'",
  start_date = start_date,
  end_date = end_date,
  region = 8371,
  region_source = "EEZ"
  )
```


```{r advanced_gfw_speed, eval = FALSE, echo = FALSE}
speed_categories <- c("<2",
                      "2-4",
                      "4-6",
                      "6-10",
                      "10-15",
                      "15-25",
                      ">25"
                      )
sp <- paste0("speed ='" , speed_categories, "'")
vp_all_speeds <- purrr::map(sp,
                            ~gfw_ais_presence(
                              spatial_resolution = "LOW",
                              temporal_resolution = "MONTHLY",
                              group_by = "FLAGANDGEARTYPE",
                              filter_by = .x,
                              start_date = start_date,
                              end_date = end_date,
                              region = id$id, 
                              region_source = "EEZ"
                            ))
vp_all_speeds <- purrr::map2(vp_all_speeds, speed_categories, 
           ~mutate(.x, speed = .y)) |> 
  bind_rows()
vp_all_speeds <- vp_all_speeds |> mutate(speed = as.factor(speed))
vp_all_speeds |>   
ggplot() +
  geom_tile(aes(x = Lon,
                y = Lat,
                fill = `Vessel Presence Hours`)) +
  geom_sf(data = ne_countries(returnclass = "sf", scale = "medium")) +
  coord_sf(xlim = c(min(eez_vessel_presence_speed$Lon),max(eez_vessel_presence_speed$Lon)),
           ylim = c(min(eez_vessel_presence_speed$Lat),max(eez_vessel_presence_speed$Lat))) +
  scale_fill_gradientn(
    trans = 'log10',
    colors = map_effort_light, 
    na.value = NA,
    labels = scales::comma) +
  facet_wrap(peed) +
  labs(title = "Vessel Presence hours in the Senegalese EEZ",
       subtitle = glue("{start_date} to {end_date}"),
       fill = "Fishing hours") +
  map_theme
```
-->

## Mapping vessel presence with ggplot2

Before mapping let's define a theme using `ggplot2`

```{r theme}
# Map theme with dark background
map_theme <- ggplot2::theme_minimal() + 
  ggplot2::theme(
    panel.border = element_blank(), 
    legend.position = "bottom", legend.box = "vertical", 
    legend.key.height = unit(3, "mm"), 
    legend.key.width = unit(20, "mm"),
    legend.text = element_text(color = "#848b9b", size = 8), 
    legend.title = element_text(face = "bold", color = "#363c4c", size = 8, hjust = 0.5), 
    plot.title = element_text(face = "bold", color = "#363c4c", size = 10), 
    plot.subtitle = element_text(color = "#363c4c", size = 10), 
    axis.title = element_blank(), 
    axis.text = element_text(color = "#848b9b", size = 6)
    )

# Palette for fishing activity
map_effort_light <- c("#ffffff", "#eeff00", "#3b9088","#0c276c")
```


And let's map the original vessel presence dataset for January-March 2024: 


```{r eez_df}
vp_senegal
```


We can use `ggplot2` and `geom_tile` to plot the data.

```{r sen_plot, fig.width=8, fig.height=6}
#| fig.alt: >
#|   A map of Vessel Presence in the Senegalese EEZ. 
#|   The scale is logarithmic
vp_senegal |> 
  ggplot() +
  geom_tile(aes(x = Lon,
                y = Lat,
                fill = `Vessel Presence Hours`)) +
  geom_sf(data = ne_countries(returnclass = "sf", scale = "medium")) +
  coord_sf(xlim = c(min(vp_senegal$Lon), max(vp_senegal$Lon)),
           ylim = c(min(vp_senegal$Lat), max(vp_senegal$Lat))) +
  scale_fill_gradientn(
    trans = 'log10',
    colors = map_effort_light, 
    na.value = NA,
    labels = scales::comma) +
  labs(title = "Vessel Presence hours in the Senegalese EEZ",
       subtitle = glue("{start_date} to {end_date}"),
       fill = "Vessel presence hours (log)") +
  map_theme
```



