---
title: "Using get_vessel_info(): the basics of vessel identity in Global Fishing Watch and gfwr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using get_vessel_info(): the basics of vessel identity in Global Fishing Watch and gfwr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#",
  eval = TRUE
)
```

```{r, eval = F}
library(gfwr)
```

```{r load, eval = T, include = FALSE}
devtools::load_all() 
```

```{r, message=FALSE}
library(dplyr)
library(tidyr)
```

This vignette explains the use of `get_vessel_info()` as a key function to 
understand vessel identity and to use all the other Global Fishing Watch API 
endpoints.

We discuss basic identity markers, why Vessel ID was created, the structure of 
the `get_vessel_info()` response and how to get __`vesselId`__ for use in the 
functions `get_event()` and `get_event_stats()`. 


## Automatic Identification System (AIS) 

The Automatic Identification System (AIS) is an automatic tracking system 
originally developed to help preventing collisions between vessels at sea. 
Vessels broadcast AIS to  alert other vessels of their presence, but terrestrial
and satellite receivers can also receive these messages and monitor vessel 
movements. 
AIS is at the core of Global Fishing Watch analysis pipelines, including the 
AIS-based fishing effort calculation displayed on our map and available through 
function `get_raster()` in `gfwr`.

AIS messages include identity information about the vessel, such as ship name, 
call sign, and International Maritime Organization (IMO) number, as well as an 
identifier known as the Maritime Mobile Service Identity (MMSI). 

- __MMSI__ are nine-digit numbers broadcasted in AIS messages. MMSIs are 
supposed to be unique for each vessel, but 1. a vessel can change MMSIs 
throughout its lifecycle--for example when it's reflagged, because the first 
three digits refer to the flag country 2. several vessels can broadcast the same
MMSI at the same time. This happens for many reasons, including the fact that 
data entry in AIS messages is manual.

- __Shipname__ and __callsign__ can be also transmitted in AIS messages but they
are optional, not every AIS-broadcasting vessel transmits them, and their
transmission can be inconsistent. Shipnames can also vary a lot in their 
spelling, requiring some fuzzy matching to detect if they refer to the same 
vessel. 

- __IMO numbers__ are permanent unique identifiers that follow a vessel from 
construction to scrapping. Assigned by the International Maritime Organization, 
IMO numbers are required for only a subset of industrial fishing vessels. IMO 
number can be transmitted along with MMSI in AIS messages but they are 
frequently missing.

These identity markers are often the starting point of any inquiry around vessel
identity.
However, due to their characteristics, none of these identifiers should be 
interpreted as the sole identity source for a vessel. Global Fishing Watch does 
extensive work to analyze and gather all the information available for a given 
vessel into cohesive sets. 

> __Note:__
MMSI is referred to as `ssvid` in our tables. `ssvid` stands for 
"source-specific vessel identity". In this case, the source is AIS, and 
ssvid = MMSI.


## __`vesselId`__

To solve the complexity of having several vessel identifiers that can be 
duplicated or missing for each vessel and that can change in time, Global 
Fishing Watch developed __`vesselId`__, a vessel identity variable that 
combines vessel information and is specific to a specific time interval.


A __`vesselId`__ is formed by a combination of the MMSI and the IMO number when 
available, or by the MMSI, callsign and shipname transmitted in AIS messages. 
Each __`vesselId`__ is associated to a single MMSI at a specific period of 
time, and refers to a single vessel.

On the other side, a single vessel can have several __`vesselId`__ in time, and 
this is why simple calls to `get_vessel_info()` can return tables that have many
`vesselIds` and different identity markers in time.


## Basic and advanced searches

The function `get_vessel_info()` allows a user to run a basic query using MMSI, 
callsign, shipname or IMO number but it also allows for complex searches, using 
a combination of these to retrieve the vessel of interest more accurately:

__Do a simple search using "query" and search_type = "search" (which is the 
default so it can be omitted from the function call)__

```{r search, eval = FALSE}
get_vessel_info(query = 224224000, search_type = "search")
```

__Do complex search or fuzzy matching using `"where"` and `search_type = "search"`__

```{r, eval= FALSE}
get_vessel_info(where = "imo = '8300949'")
get_vessel_info(where = "imo = '8300949' AND ssvid = '214182732'")
get_vessel_info(where = "shipname LIKE '%GABU REEFE%' OR imo = '8300949'")
```


Importantly, __the response will return all the information it has for the vessel 
that matches the combination of identity markers requested, not only the ones 
requested__.

This means that the function does not "filter" the results as requested in the 
function call. Instead, the function returns all the `vesselIds` belonging to the 
same vessel.

> **Note**:
> The same logic does not apply to `get_event()`: calls to `get_event()` using a 
single `vesselId` will return events only for the requested `vesselId`.


## Use examples 

Let's go back to the simple search. To get information of a vessel using its 
MMSI, IMO number, callsign or name, the search can be done directly using the 
number or the string. For example, to look for a vessel with `MMSI = 224224000`,
the number is enough:

```{r example_vessel_info_1}
mmsi_search <- get_vessel_info(224224000)
```

The response from the API is a list with seven elements: 

```{r str}
names(mmsi_search)
```

The content of the original AIS messages transmitted by the vessel appears in 
`$selfReportedInfo`:

```{r self, max.width=10}
mmsi_search$selfReportedInfo
```

As you can see, this vessel returns a dataframe with two rows. One corresponds 
to our original search, where ssvid (MMSI) equals 224224000.

The second line has a different ssvid, but the same name and IMO number.

The two lines correspond to the same vessel, and as you can see from the fields 
`transmissionDateFrom` and `transmissionDateTo`, `flag`, and `ssvid`, the
vessel operated with a Spain flag (ESP) and one ssvid between 2015 and 2019, then
it was reflagged and operated with a BES flag (from Bonaire, Sint Eustatius and Saba) 
between 2019 and 2023. The change in ssvid reflects the reflagging operation 
because the first three digits of MMSI correspond to the flag country.

Variable `matchFields` reports that the matching was done using `"SEVERAL_FIELDS"`. 



```{r message=FALSE}
mmsi_search$selfReportedInfo %>% 
  select(vesselId, ssvid, flag, contains("Date"))
```

This is a simple case, in which the successive __`vesselId`__ do not overlap in time 
and most identifiers match, in spite of some changes.

For some vessels, variables `transmissionDateFrom` and `transmissionDateTo` can 
overlap and other fields can be different. 


### Using __`vesselId`__ in other functions

__`vesselId`__ can be extracted from `$selfReportedInfo$vesselId`, but it is highly 
recommended to take a look at the response and confirm which of the values returned 
as `vesselId` should be selected. 

Before picking a __`vesselId`__ to use in other functions, it is useful to examine:

- which __`vesselId`__ corresponds to the time interval of interest
- if other identifiers are matching, indicating there is enough confidence that 
different __`vesselId`__ refer to the vessels of interest
- the number of messages transmitted, in `messagesCounter`. Sometimes very few 
positions are transmitted for a short time interval and that __`vesselId`__ can 
be treated as an exception

You can use the selected __`vesselId`__ to get any events related to the vessel
of interest in other functions.

Let's find encounter events for all the `vesselIds` from the previous search:

```{r}
id <- mmsi_search$selfReportedInfo$vesselId
id
events <- get_event(event_type = "ENCOUNTER", vessels = id) 
events
```

In our example, Global Fishing Watch analyses report that the vessel had 
`r nrow(events)` 
encounters.

## Registry data 

Vessel registries carry important vessel identity information, like vessel
characteristics, registration history, licenses to fish in certain areas, and
vessel ownership data. Global Fishing Watch compiles vessel information from over
40 public vessel registries and matches this information with the AIS-transmitted
identity fields to provide a better overview of a vessel's identity. 

This information is requested by parameter `"includes"` and returned in the element 

- `$registryInfoTotalRecords` (number of records in registries),

```{r}
mmsi_search$registryInfoTotalRecords
```

- `$registryInfo` the actual information in the registry, including identity, 
vessel characteristics and dates of transmission

```{r}
mmsi_search$registryInfo
```

- `$registryOwners` with their name, country of origin, ssvid of the vessel and 
dates of ownership. Sometimes the vessel can change identities and flags but its 
owners remain the same, and sometimes changes in identity correspond to changes in ownership. 

```{r}
mmsi_search$registryOwners
```

- `$registryPublicAuthorizations` of the response and the respective organizations

```{r registryInfo}
mmsi_search$registryPublicAuthorizations %>%
  tidyr::unnest(sourceCode)
```

## Matching AIS and registries: case examples

In the best of cases, AIS messages match registry information and the whole identity 
of the vessel can be reconstructed. 
Here are two examples with registry and AIS data not overlapping in time.

This vessel has a single `vesselId` throughout its entire history:

```{r one AIS}
one_vesselId <- get_vessel_info(where = "ssvid='701024000'")

#see AIS-based identities
one_vesselId$selfReportedInfo
#check registry info:
one_vesselId$registryInfo %>%
  dplyr::relocate(transmissionDateFrom, transmissionDateTo) #changing column order for visualization
```

This other vessel has had more than one `vesselId` based on AIS, but the history
is easy to reconstruct:

```{r remedy002}
multiple_vesselIds <- get_vessel_info(where = "ssvid='412217989'") 
# see AIS-based identities:
multiple_vesselIds$selfReportedInfo %>%
  relocate(transmissionDateFrom, transmissionDateTo) 
#check registry info:
multiple_vesselIds$registryInfo 
```

However, sometimes a vessel found in AIS has no registry information and the registry fields come back empty. It is also possible that a search returns a 
vessel with no matching AIS information and no registry.



# Read more:

- Park et al. (2023). Tracking elusive and shifting identities of the global fishing fleet [link](https://www.science.org/doi/full/10.1126/sciadv.abp8200) for a more detailed account of Global Fishing Watch identity work.
- https://globalfishingwatch.org/datasets-and-code-vessel-identity/
- https://globalfishingwatch.org/data/spoofing-one-identity-shared-by-multiple-vessels/ 


